Syntax and Semantics of Cedille

### 2020-12-25 docs/semantics/paper.tex
docs: semantics: modify substitution for bidirectional rules

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}, with
Figure~\ref{fig:ctxt} giving the context formation rules.
For brevity, we take these figures as implicitly specifying the syntax of
contexts \(\Gamma\), kinds $\kappa$, types $T$, and annotated terms $t$; these
may use term variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
it is intended that such occurrences be read the same way (i.e., either all as
\(\tpcheck\) or all as \(\tpsynth\)).
We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
be a term).
The rules are intended, with a few points of nondeterminism, to be read
bottom-up (in a standard way;
c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
an algorithm for synthesizing a classifier from a context and an expression
($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
($\tpcheck$) (\emph{type checking}).
Specifically, nondeterminism arises in the two rules allowing both synthesized
and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
constructors required by other rules may be revealed.
The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
is sufficient to reduce types to weak head normal form.

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\    
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

@@

    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} &
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
          {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t_1 \simeq t_2\}$ for definitionally equal terms (that is,
  terms that are \(\beta\eta\)-equivalent modulo erasure), one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to (a possibly unrelated) $|t'|$.
  We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any term $t'$ as a realizer of $\{ t_1
  \simeq t_2\}$, provided the free variables of \(t'\) are declared in the
  context.
  This means that in Cedille, any such term --- even otherwise untypable
  terms, non-normalizing terms, etc. --- prove trivially true equations.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2\) in the synthesized type of the second subexpression.
  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
  indicates the occurrences of \(t_1\) to substitute.
  \([t_2/x]T\) must be a well-kinded type, but it is not required that
  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
  type, or will even be typeable.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic is used to produce a resulting type that
  is well-kinded whose details are beyond the scope of this document.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\ \{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

>>>>

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}, with
Figure~\ref{fig:ctxt} giving the context formation rules.
For brevity, we take these figures as implicitly specifying the syntax of
contexts \(\Gamma\), kinds $\kappa$, types $T$, and annotated terms $t$; these
may use term variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
it is intended that such occurrences be read the same way (i.e., either all as
\(\tpcheck\) or all as \(\tpsynth\)).
Since term variables synthesize their types but the usual notion of substitution may
replace these with terms whose types can only be checked, we introduce a
short-hand notation: let \([t/x]^T\) stand for \([\chi\ T\ \mhyph\ t/x]\),
where the \(\chi\) construct gives an explicit type annotation (see
Section~\ref{sec:overview}).

We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
be a term).
The rules are intended, with a few points of nondeterminism, to be read
bottom-up (in a standard way;
c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
an algorithm for synthesizing a classifier from a context and an expression
($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
($\tpcheck$) (\emph{type checking}).
Specifically, nondeterminism arises in the two rules allowing both synthesized
and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
constructors required by other rules may be revealed.
The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
is sufficient to reduce types to weak head normal form.

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

\begin{figure}
  \[
    \begin{array}{c}
  \begin{array}{ll}
    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\    
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

@@

    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} &
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]^TT' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]^TT'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
          {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
      \\ \\
      \texttt{Top} = \{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x} \}
    \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t_1 \simeq t_2\}$ for definitionally equal terms (that is,
  terms that are \(\beta\eta\)-equivalent modulo erasure), one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to (a possibly unrelated) $|t'|$.
  We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any term $t'$ as a realizer of $\{ t_1
  \simeq t_2\}$, provided the free variables of \(t'\) are declared in the
  context.
  This means that in Cedille, any such term --- even otherwise untypable
  terms, non-normalizing terms, etc. --- prove trivially true equations.
  Put another way, any trivially true equation type in CDLE is a suitable type
  to classify all untyped lambda calculus terms.
  We define such a type \texttt{Top} at the bottom of Figure~\ref{fig:tp}, which
  is used in the typing of the next construct we discuss, \(\rho\).
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2\) in the synthesized type of the second subexpression.
  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
  indicates the occurrences of \(t_1\) to substitute.
  \([t_2/x]^{\texttt{Top}}\ T\) must be a well-kinded type, but it is not required that
  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
  type, or will even be typeable.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic is used to produce a resulting type that
  is well-kinded whose details are beyond the scope of this document.

  It might appear at first that the requirement that the guide \(T\) be written
  such that the resulting substitution is well-kinded is too strict.
  For example, what if we are trying to rewrite \(n\) in \(X\ n\), where \(X\)
  is a predicate on (Church-encoded) naturals?
  The next construct, \(\varphi\), gives us the ability to recover the typing
  information associated to \(n\).
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\ \{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.
